---
title: "Large-lot zoning is a relic of Jim Crow. It's time to abolish it."
author: "pete rodrigue"
date: "2025-12-04"
output: 
  html_document:
    self_contained: true
---

<style type="text/css">
body{
  font-size: 14pt;
}
</style>

<br>

This brief examines how the Federal Housing Administration's explicitly racist neighborhood grades from the 1930s intersect with DC's current zoning policies and socioeconomic inequality. "A" and "B" grades designated the wealthiest white neighborhoods, while lower grades (down to "H") marked moderate-to-lower-income, racially diverse, or Black neighborhoods.  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning=F, message=F, out.width = "100%")

# should appear at
# https://raw.githack.com/pete-rodrigue/hd_analysis/main/exclusionary-zoning-analysis.html

# load packages
library(dplyr)
library(readr)
library(sf)
library(terra)
library(leaflet)
library(leaflet.extras)
# devtools::install_github("statnmap/HatchedPolygons")
library(HatchedPolygons)
# install.packages('eulerr')
library(eulerr)
library(ggplot2)
library(patchwork)

# Getting the path of your current open file
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))
# load functions we've written to help with the analysis below
# (we've stored those in a separate file to make the code easier to read)
source("supporting_functions.R")

# define colors
abc_col = "#E8B391"
fgh_col = "#B15B25"
r1_col  = "#25B15B"
```

```{r load_data}
####################
# LOAD BLOCK DATA
####################
# Load raw block data:
# 1950 and 1960 block data from here: https://blogs.gwu.edu/centerforwashingtonareastudies/resources/. Compiled by Leah Brooks and team.
b40_shp <- sf::st_read("block_shapes/1940_block_shapefiles/1940_census_DC_joined.shp", quiet=T)
b50_shp <- sf::st_read("block_shapes/1950_block_shapefiles/1950_census_blocks_joined.shp", quiet=T)
b60_shp <- sf::st_read("block_shapes/1960_block_shapefiles/1960_census_blocks_joined.shp", quiet=T)
# other block data come from IPUMS NHGIS
b70_shp <- sf::st_read("block_shapes/DC_block_1970/DC_block_1970.shp", quiet=T)
b80_shp <- sf::st_read("block_shapes/DC_block_1980/DC_block_1980.shp", quiet=T)
b90_shp <- sf::st_read("block_shapes/nhgis0092_shapefile_tl2000_110_block_1990/DC_block_1990.shp", quiet=T)
b00_shp <- sf::st_read("block_shapes/nhgis0092_shapefile_tl2000_110_block_2000/DC_block_2000.shp", quiet=T)
b10_shp <- sf::st_read("block_shapes/nhgis0092_shapefile_tl2010_110_block_2010/DC_block_2010.shp", quiet=T)
b20_shp <- sf::st_read("block_shapes/nhgis0092_shapefile_tl2020_110_block_2020/DC_block_2020.shp", quiet=T)

b70_df <- readr::read_csv("block_data/nhgis0093_ds96_1970_block.csv")
b80_df <- readr::read_csv("block_data/nhgis_ds104_1980_block_11.csv")
b90_df <- readr::read_csv("block_data/nhgis0092_ds120_1990_block.csv")
b00_df <- readr::read_csv("block_data/nhgis0092_ds147_2000_block.csv")
b10_df <- readr::read_csv("block_data/nhgis0092_ds172_2010_block.csv")
b20_df <- readr::read_csv("block_data/nhgis0092_ds258_2020_block.csv")

invisible(gc())

# rework 1940, 1950 and 1960 data a little bit:
b40_shp <- 
  b40_shp %>% 
  mutate(n_tot = as.numeric(tot_occ), 
         n_white = as.numeric(tot_occ) - as.numeric(nw_occ),
         n_black = n_tot - n_white) %>% 
  select(n_tot, n_black, n_white) %>% 
  mutate(geo_id = as.character(row_number())) %>% mutate(year=1940) %>% 
  sf::st_transform(., 26918) %>%
  mutate(geo_area_meters = sf::st_area(.))
b50_shp <- 
  b50_shp %>% 
  mutate(n_tot = as.numeric(total_occ), 
         n_white = as.numeric(total_occ) - as.numeric(non_white_),
         n_black = n_tot - n_white) %>% 
  select(n_tot, n_black, n_white) %>% 
  mutate(geo_id = as.character(row_number())) %>% mutate(year=1950) %>% 
  sf::st_transform(., 26918) %>%
  mutate(geo_area_meters = sf::st_area(.))

b60_shp <- 
  b60_shp %>% 
  # the 1960 data doesn't have race cross tabs by individual person,
  # but we can take the household race breakdown and apply it to the 
  # population. this is probably roughly accurate. 
  # as a robustness check we can just use the household data and rerun
  # the analysis; see the commented out code.
  mutate(n_tot = as.numeric(stringr::str_remove_all(total_pop, "\\*")), 
         pct_white = (as.numeric(total_occ) - as.numeric(nonwhite_o)) / as.numeric(total_occ)) %>% 
  # mutate(n_tot = as.numeric(total_occ), 
  #        n_white = as.numeric(total_occ) - as.numeric(nonwhite_o)) %>% 
  mutate(n_white = pct_white * n_tot,
         n_black = n_tot - n_white) %>%
  select(n_tot, n_black, n_white) %>% 
  mutate(geo_id = as.character(row_number())) %>% mutate(year=1960) %>% 
  sf::st_transform(., 26918) %>%
  mutate(geo_area_meters = sf::st_area(.))

# 1970 block data has to be specially cleaned, see https://forum.ipums.org/t/race-ethnicity-data-at-a-block-level-from-1970/6178
b70_df$c_black <- b70_df$CM6001 + b70_df$CM6002
b70_df$c_other <- b70_df$CM6003 + b70_df$CM6004
b70_df$c_white <- b70_df$CM5001 + b70_df$CM5002 - b70_df$c_black - b70_df$c_other

# merge the data and the shapefiles and clean up the data:
b70_shp <- clean_block_data(b70_shp, b70_df, "GISJOIN", "GISJOIN", "c_", "c_black", "c_white", year=1970)
b80_shp <- clean_block_data(b80_shp, b80_df, "GISJOIN", "GISJOIN", "C9D0", "C9D002", "C9D001", year=1980)
b90_shp <- clean_block_data(b90_shp, b90_df, "GISJOIN", "GISJOIN", "EUY0", "EUY002", "EUY001", year=1990)
b00_shp <- clean_block_data(b00_shp, b00_df, "GISJOIN", "GISJOIN", "FYE0", "FYE002", "FYE001", year=2000)
b10_shp <- clean_block_data(b10_shp, b10_df, "GISJOIN", "GISJOIN", "H7X", "H7X003", "H7X002", "H7X001", year=2010)
b20_shp <- clean_block_data(b20_shp, b20_df, "GISJOIN", "GISJOIN", "U7J", "U7J003", "U7J002", "U7J001", year=2020)

# remove data we don't need anymore to keep our workspace neat
rm(b70_df, b80_df, b90_df, b00_df, b10_df, b20_df)
invisible(gc())

# fix any broken geometries:
b40_shp <- fix_geo_if_broken(b40_shp)
b50_shp <- fix_geo_if_broken(b50_shp)
b60_shp <- fix_geo_if_broken(b60_shp)
b70_shp <- fix_geo_if_broken(b70_shp)
b80_shp <- fix_geo_if_broken(b80_shp)
b90_shp <- fix_geo_if_broken(b90_shp)
b10_shp <- fix_geo_if_broken(b10_shp)
b20_shp <- fix_geo_if_broken(b20_shp)

# combine shapes into one big object, remove individual objects:
geos_shp <- dplyr::bind_rows(b40_shp, b50_shp, b60_shp, b70_shp, b80_shp, 
                             b90_shp, b00_shp, b10_shp, b20_shp)
# create a truly unique block/tract ID:
geos_shp$geo_id <- paste0(geos_shp$year, "_", geos_shp$geo_id)

# remove objects we don't need anymore
rm(list=ls(pattern="^b[0-9]{2}"))
rm(list=ls(pattern="^t[0-9]{2}"))
# garbage collection
invisible(gc())


####################
# LOAD COVENANT DATA
####################
dc_dcs_shp <- sf::st_read("discriminatory_policies/dc_discriminatory_covenants/dc_discriminatory_covenants.shp", quiet=T)

dc_dcs_shp_detail <-
  filter(dc_dcs_shp, !(fid %in% c(265, 266, 236, 206, 186, 141, 173, 231, 234, 226, 205, 234, 242, 225, 205, 144, 124, 136, 139)))

# remove some shapes that are just artifacts (in PG county)
dc_dcs_shp <-
  filter(dc_dcs_shp, !(fid %in% c(265, 266, 236, 206, 186, 141, 173, 231, 234, 226, 205, 234, 242, 225, 205, 144, 124, 136, 139))) %>%
  sf::st_simplify(., dTolerance = .00002)


# uncomment to look at covenants data on a map
# leaflet() %>%
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolygons(data=dc_dcs_shp_detail %>% st_transform(4326),
#               fillColor = "#F4B942",
#               color = "#F4B942",
#               stroke=T,
#               fillOpacity=1,
#               group="Approx. areas w/ discriminatory covenants") %>%
#   addLayersControl(
#     overlayGroups = c("Approx. areas w/ discriminatory covenants"),
#     options = layersControlOptions(collapsed = FALSE)
#   )


####################
# LOAD ZONE DATA
####################
# load and simplify the zoning map:
# comes from here: https://opendata.dc.gov/datasets/DCGIS::zoning-boundaries-zoning-regulations-of-2016/about
zone_shp <- sf::st_read("zoning/Zoning_Boundaries_(Zoning_Regulations_of_2016).shp", quiet=T)

# list zones in the ZR16 data:
zones_list <- sort(unique(zone_shp$ZR16))
housing_zones <- zones_list[grep(x=zones_list, pattern = "^R|^MU|^NMU")]
# get unzoned areas:
unzoned_shp <- filter(zone_shp, ZR16=='UNZONED')
# subset zones to mostly-housing zones 
zone_shp <- zone_shp[zone_shp$ZR16 %in% housing_zones,]

# create simplified labels
zone_shp$ZR16_simple <- "Other"
zone_shp$ZR16_simple[grep(x=zone_shp$ZR16, pattern="^RA-")] <- "Apartment zones"
zone_shp$ZR16_simple[grep(x=zone_shp$ZR16, pattern="^R-")] <- "Residential zones"
zone_shp$ZR16_simple[grep(x=zone_shp$ZR16, pattern="^RF-")] <- "Residential flat zones"
zone_shp$ZR16_simple[grep(x=zone_shp$ZR16, pattern="^MU-")] <- "Mixed use zones"
zone_shp$ZR16_simple[grep(x=zone_shp$ZR16, pattern="^NMU-")] <- "Mixed use zones"

zone_shp <- sf::st_transform(zone_shp, 26918)

# note r-1 zones
zone_shp$r_1 <- 0
zone_shp$r_1[zone_shp$ZR16 %in% grep(pattern = "R-1", x = sort(unique(zone_shp$ZR16)), value=T)] <- 1

# load the 2022 ward shape files
# comes from here: https://opendata.dc.gov/datasets/DCGIS::wards-from-2022/about
ward_shp <- 
  sf::st_read("Wards_from_2022/Wards_from_2022.shp", quiet=T) %>%
  sf::st_transform(26918)

####################
# LOAD FHA DATA
####################
fha_raster <- terra::rast("discriminatory_policies/fha_map/fha_image.tif")

fha_shp <- sf::st_read("discriminatory_policies/fha_map/fha_map.shp", quiet=T) %>% 
  mutate(fha_grade = substr(fha_grade, 1, 1)) %>%
  sf::st_transform(4326)
fha_centriods <- sf::st_centroid(fha_shp)
fha_centriods$lon <- sf::st_coordinates(fha_centriods)[,1]
fha_centriods$lat <- sf::st_coordinates(fha_centriods)[,2]

fha_shp <- st_transform(fha_shp, 26918)
race_shp <- 
  geos_shp %>% 
  filter(year==1940) %>% 
  st_transform(4326) %>% 
  mutate(n_nonwhite = n_tot - n_white,
         pct_nonwhite = n_nonwhite / n_tot)



####################
# LOAD PROPERTY DATA
####################
# CAMA data:
# https://opendata.dc.gov/datasets/DCGIS::computer-assisted-mass-appraisal-residential/about
cama <- readr::read_csv("Computer_Assisted_Mass_Appraisal_-_Residential.csv", 
                        show_col_types = F)
# Record lots data:
# https://opendata.dc.gov/datasets/5b0b6b13ef894b8da62e6bd458d907b3_35/explore?location=38.896842%2C-77.002937%2C18.31
rl <- sf::st_read("Record_Lots/Record_Lots.shp", quiet=T)

# convert property data to points (simpler, less memory)
pd <- 
  dplyr::left_join(rl, cama, by="SSL") %>% 
  filter(!is.na(OBJECTID.y)) %>%
  sf::st_centroid() %>%
  sf::st_transform(26918)

# get properties in R-1 zones
pd$in_r <- 
  lengths(
    sf::st_intersects(pd,
                      filter(zone_shp, 
                             ZR16 %in% grep(pattern = "R-1",
                                            x = sort(unique(zone_shp$ZR16)), 
                                            value=T)
                             )
                      )
    )

# get properties in former FHA grade areas:
pd$in_A <- lengths(sf::st_intersects(pd, fha_shp[fha_shp$fha_grade=="A",] %>% st_transform(st_crs(pd))))
pd$in_B <- lengths(sf::st_intersects(pd, fha_shp[fha_shp$fha_grade=="B",] %>% st_transform(st_crs(pd))))
# pd$in_C <- lengths(sf::st_intersects(pd, fha_shp[fha_shp$fha_grade=="C",] %>% st_transform(st_crs(pd))))
# 
# pd$in_F <- lengths(sf::st_intersects(pd, fha_shp[fha_shp$fha_grade=="F",] %>% st_transform(st_crs(pd))))
pd$in_G <- lengths(sf::st_intersects(pd, fha_shp[fha_shp$fha_grade=="G",] %>% st_transform(st_crs(pd))))
pd$in_H <- lengths(sf::st_intersects(pd, fha_shp[fha_shp$fha_grade=="H",] %>% st_transform(st_crs(pd))))

# get properties in wards 2 and 3:
wards_2_3 <- ward_shp %>% filter(WARD %in% c(2, 3)) %>% sf::st_union()
ward_3 <- ward_shp %>% filter(WARD %in% c(3)) %>% sf::st_union()
pd$in_ward_2_3 <- lengths(st_intersects(pd, wards_2_3))
pd$in_ward_3 <- lengths(st_intersects(pd, ward_3))

# function to get the number of houses in these zones, average value, and aggregate value:
get_stats <- function(df, rv=F) {
  n <-
    df %>%
    sf::st_drop_geometry() %>%
    summarise(n = n()) %>%
    pull(n)
  
  n <- round(n / 100, 0)*100
  
  avg_p <-
    df %>%
    sf::st_drop_geometry() %>%
    mutate(sale_date = as.numeric(substr(x = SALEDATE, start = 1, stop = 4))) %>%
    filter(sale_date >= 2020) %>%
    summarise(average_value = mean(PRICE, na.rm=T)) %>%
    pull(average_value)
  
  agg_val <- n * avg_p
  
  cat(
    paste0(
      "Number of houses: ~", format(n, big.mark=","),
      "\nAverage sale price after 2019: ~", format(avg_p, big.mark=","),
      "\nEstimated aggregate value: ~", format(agg_val, big.mark=",")
    )
  )
  
  if (rv) {return(list('n'=n, 'avg_p'=avg_p, "agg_val"=agg_val))}
}

# cat("\n---A grade area stats---\n")
# get_stats(df = pd[pd$in_A==1,])
# cat("\n\n")
# cat("---H grade area stats---\n")
# get_stats(df = pd[pd$in_H==1,])
```


The map below shows the overlap between former FHA grades and current average home prices. Also shown:

* the approximate locations of homes that had discriminatory covenants
* 1940s Black residential areas
* R-1 zones, where DC bans multifamily housing like duplexes and apartments today

Use the checkboxes to toggle layers; click and drag to navigate.

<br>


```{r big_map, out.width="100%"}
####################
# MAP EVERYTHING
####################
pdheat <- 
  bind_cols(pd, sf::st_coordinates(pd %>% sf::st_transform(4326))) %>%
  filter(PRICE > 0) %>% filter(!is.na(PRICE)) %>% 
  mutate(sale_date = as.numeric(substr(x = SALEDATE, start = 1, stop = 4))) %>%
  filter(sale_date >= 2010) %>%
  filter(PRICE >= 100000) %>%
  select(X, Y, PRICE) %>%
  mutate(log_price = log10(PRICE)  )

pdgrp <- st_join(geos_shp %>% filter(year==2020) %>% select(geo_id),
                 pdheat, 
                 join = st_intersects) %>%
  group_by(geo_id) %>%
  summarize(price = mean(PRICE, na.rm=T),
            log_price = log10(price))

pdgrp <-
  pdgrp %>%
  filter(!is.na(price)) %>% filter(!is.nan(price))

fha_pal <- colorFactor(palette = "RdYlGn", domain = fha_shp$fha_grade, reverse = T)

race_pal <- colorNumeric(palette = "YlOrBr", domain = race_shp$pct_nonwhite)

price_pal <- colorNumeric(palette = viridisLite::viridis(n = 256, option = "A") , domain = log10(pdgrp$price))

zone_hatched <-
  zone_shp %>%
  filter(ZR16 %in% grep(pattern = "R-1", x = sort(unique(zone_shp$ZR16)), value=T)) %>%
  st_transform(4326) %>%
  hatched.SpatialPolygons(., density = 500)



leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data=fha_shp %>% st_transform(4326),
              fillColor = ~fha_pal(fha_grade),
              stroke=F,
              fillOpacity = .8,
              label=~paste("grade:", fha_grade),
              group="FHA grades from the 30s") %>%
  addPolygons(data=race_shp,
              fillColor = ~race_pal(pct_nonwhite),
              stroke=F,
              fillOpacity = ~ifelse(is.na(pct_nonwhite), 0, 0.8),
              label=~paste0(round(pct_nonwhite*100, 0), "% non-white residents in 1940"),
              group="% non-white in 1940") %>%
  # addLabelOnlyMarkers(lng = fha_centriods$lon, lat = fha_centriods$lat, group = "FHA grade",
  #                     label = fha_centriods$fha_grade, labelOptions = c(permanent=T)) %>%
  addPolygons(data=dc_dcs_shp %>% st_transform(4326),
              fillColor = "#F4B942",
              color = "#F4B942",
              stroke=T,
              fillOpacity=1,
              group="Approx. areas w/ discriminatory covenants") %>%
  addRasterImage(fha_raster, opacity = 0.7, group="Paper FHA map") %>% # Add raster with desired opacity
  addPolygons(data=zone_hatched,
              fillColor = "#9ffcb1",
              color = "#9ffcb1",
              fillOpacity = .6,
              opacity=.6,
              group="Today's large-lot zones 'R-1x'") %>%
  addPolygons(data=pdgrp %>% st_transform(4326),
              group="Recent home prices",
              fillColor = ~price_pal(log_price),
              stroke=F,
              fillOpacity = .95,
              label=~format(round(price, 0), big.mark=",")
              ) %>%
  addLayersControl(
    overlayGroups = c("Approx. areas w/ discriminatory covenants", "% non-white in 1940",
                      "FHA grades from the 30s", 
                      # "FHA grade", 
                      "Paper FHA map",
                      "Recent home prices",
                      "Today's large-lot zones 'R-1x'"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  addLegend(
    position = "bottomright",
    colors = "#9ffcb1",
    labels = "",
    title = "Current zones that prohibit<br>apartments and duplexes",
    opacity = 1
  ) %>%
  addLegend(
    position = "bottomright",  # Or "topright", "bottomleft", "topleft"
    pal = fha_pal,
    values = fha_shp$fha_grade,
    title = 'FHA "grade"',
    opacity = 1
  ) %>%
  hideGroup("Approx. areas w/ discriminatory covenants") %>%
  hideGroup("FHA map") %>%
  hideGroup("% non-white in 1940") %>%
  hideGroup("Paper FHA map") %>%
  hideGroup("Recent home prices") %>%
  hideGroup("Today's large-lot zones 'R-1x'")
```



```{r disparities, out.width="100%"}
temp <-
  sf::st_join(fha_shp %>% filter(fha_grade %in% c("A", "B", 
                                                  # "C", "F", 
                                                  "G", "H")),  
              geos_shp %>% filter(year==2020) %>% sf::st_centroid()) %>%
  sf::st_drop_geometry() %>%
  mutate(geo_area_meters = as.vector(geo_area_meters)) %>%
  group_by(fha_grade) %>%
  summarise(
    n_tot          = sum(n_tot, na.rm=T),
    n_black        = sum(n_black, na.rm=T),
    n_white        = sum(n_white, na.rm=T),
    geo_area_acres = sum(geo_area_meters, na.rm=T) / 4046.86
  ) %>%
  mutate(
    pop_density    = n_tot / geo_area_acres,
    pct_white      = round(n_white / n_tot * 100, 0),
    pct_black      = round(n_black / n_tot * 100, 0)
  )

temp$color <- NA
temp$color[temp$fha_grade %in% c("A", "B", "C")] <- abc_col
temp$color[temp$fha_grade %in% c("F", "G", "H")] <- fgh_col

p1 <-
  ggplot(temp, aes(y=pct_white, x= fha_grade)) +
  geom_bar(aes( fill=abc_col), stat="identity") +
  geom_text(aes(label = format(pct_white, big.mark=",")), vjust = -0.5) +
  scale_fill_identity() +
  theme_minimal() +
  xlab("% white residents") +
  ylab("") +
  ylim(c(0, 100)) +
  # ggtitle("% of white residents in 2020 lower in former FHA grade G & H areas") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

p2 <-
  ggplot(temp, aes(y=pct_black, x= fha_grade)) +
  geom_bar(aes( fill=fgh_col), stat="identity") +
  geom_text(aes(label = format(pct_black, big.mark=",")), vjust = -0.5) +
  scale_fill_identity() +
  theme_minimal() +
  xlab("% black residents") +
  ylab("") +
  ylim(c(0, 100)) +
  # ggtitle("% of black residents in 2020 higher in former FHA grade G & H areas") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

temp <-
  pd %>%
  sf::st_drop_geometry() %>%
  filter(PRICE > 0) %>% filter(!is.na(PRICE)) %>% 
  mutate(sale_date = as.numeric(substr(x = SALEDATE, start = 1, stop = 4))) %>%
  filter(sale_date >= 2019) %>%
  filter(PRICE >= 100000) %>%
  select(PRICE, starts_with("in_")) %>%
  mutate(log_price = log10(PRICE)  ) %>%
  mutate(
    fha_grade = case_when(
      in_A == 1 ~ "A",
      in_B == 1 ~ "B",
      # in_C == 1 ~ "C",
      # in_F == 1 ~ "F",
      in_G == 1 ~ "G",
      in_H == 1 ~ "H"
    )
  ) %>%
  filter(!is.na(fha_grade)) %>%
  group_by(fha_grade) %>%
  summarise(n = n(),
            avg_p = round(mean(PRICE, na.rm=T), 0),
            agg_p = round(n*avg_p, 0))

p3 <-
  ggplot(temp, aes(y=avg_p, x= fha_grade)) +
  geom_bar(aes( fill="skyblue"), stat="identity") +
  geom_text(aes(label = paste0(round(avg_p/ 1000000, 1), "M")), vjust = -0.5, size=3) +
  scale_fill_identity() +
  theme_minimal() +
  xlab("Avg home sale $") +
  ylab("") +
  # ggtitle("Avg. home sale price correlated strongly with former FHA grade") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())
```


<br>

Today, homes in the areas the FHA reserved for rich white people (grades "A" and "B") remain disproportionately white, while "G" and "H" areas are nearly 90% black. Homes in "A" areas sell for 4 times more than homes in the lowest graded areas.

<br>


```{r plot_bars}
(p1 | p2 | p3) +
  plot_annotation(title = "Former FHA 'whites only' areas remain disproportionately rich & white today") & # Add title
  theme(plot.title = element_text(hjust = 0.5)) # Center the title
```

```{r make_venn_diagrams, out.width="100%"}
input = 1 / 4046.86 * c(
  "grade A"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="A",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  "grade B"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="B",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  # "grade C"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="C",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  # "grade D"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="D",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  # "grade E"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="E",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  # "grade F"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="F",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  "grade G"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="G",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  "grade H"=sum(as.vector(sf::st_area(st_union(st_difference(st_union(fha_shp[fha_shp$fha_grade=="H",]), st_union(zone_shp[zone_shp$r_1==1,])))))),
  
  "R-1"        =sum(as.vector(sf::st_area(st_union(st_difference(st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade %in% c("A", "B", 
                                                                                                                                                 # 'C',  'D', 'E', 'F', 
                                                                                                                                                 'G', "H"),])))))),
  
  "grade A&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="A",]))))),
  "grade B&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="B",]))))),
  # "grade C&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="C",]))))),
  # "grade D&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="D",]))))),
  # "grade E&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="E",]))))),
  # "grade F&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="F",]))))),
  "grade G&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="G",]))))),
  "grade H&R-1"=sum(as.vector(st_area(sf::st_intersection(x = st_union(zone_shp[zone_shp$r_1==1,]), st_union(fha_shp[fha_shp$fha_grade=="H",])))))
) 




temp <- data.frame(input) 
temp$grade <- rownames(temp)
temp <- temp[grepl("&", temp$grade), ]
rownames(temp) <- NULL
temp$grade <- sub("&.*", "", temp$grade)
temp$color <- NA
temp$color[temp$grade %in% c("grade A", "grade B", "grade C")] <- abc_col
temp$color[temp$grade %in% c("grade F", "grade G", "grade H")] <- fgh_col

temp$grade <- gsub(pattern = "grade ", replacement = "", x = temp$grade)
  
# temp$grade <- factor(temp$grade, levels = c("A", "B", "C", "F", "G", "H"))

p1 <- ggplot(temp, aes(y=input, x= grade)) +
  geom_bar(aes( fill=color), stat="identity") +
  geom_text(aes(label = format(round(input, 0), big.mark=",")), vjust = -0.5) +
  scale_fill_identity() +
  theme_minimal() +
  xlab("") +
  ylab("Acres of R-1 zone") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

p2 <- 
  plot(eulerr::euler(input, shape = "ellipse"),
     fills = c(abc_col, abc_col, 
               # abc_col, fgh_col, 
               fgh_col, fgh_col, r1_col)
     )

combined_patchwork <- (patchwork::wrap_elements(p1) | patchwork::wrap_elements(p2)) +
  plot_annotation(title = "OP has overwhelmingly sited exclusionary R-1 zones in former FHA 'whites only' areas") & # Add title
  theme(plot.title = element_text(hjust = 0.5)) # Center the title
```


<br>

DC's Office of Planning and Zoning Commission perpetuate this segregation by allowing rich, predominantly white Northwest DC residents to use R-1 zoning to block more affordable housing types like duplexes, fourplexes, and apartments.

Most "A" and "B" grade areas are R-1 zoned today, while few "G" and "H" areas receive this designation.

<br>

```{r plot_venn}
combined_patchwork + plot_layout(widths = c(1, 1.25), heights = c(1, 1))
```

<br>

Sociologists, historians, and economists have known for decades that "single-family" zones, like DC's R-1 zone, are used by the rich to further socioeconomic segregation.

Here's [Eleanor Noble](https://www.urban.org/sites/default/files/2022-09/Understanding%20Historic%20Exclusion_Noble%20Testimony.pdf) of the Urban Institute speaking to the DC Zoning Commission in 2022 (emphasis added):

>Zoning and land-use policies have played a central role in exacerbating racial economic and housing disparities. **Although explicit race-based zoning was struck down by the Supreme Court in 1917, racist land-use and zoning policies continued in DC through redlining, racial covenants, apartment bans and large-lot minimums, and exclusionary single-family zoning.**

Here's [Yesim Sayin](https://www.dcpolicycenter.org/publications/single-family-zoning-2019/), the director of the DC Policy Center, in 2019 (emphasis added):

>**[D]oes the District need to change single-family zoning? Yes, if the city wants to become more affordable and inclusive.** Data show us that income distribution in the city and racial composition of its neighborhoods closely follow single-family zoning... **legislators should focus on reforming the entire zoning code—and especially zoning for single-family units[.]**

Here's [Jenny Schuetz](https://www.brookings.edu/articles/charts-of-the-week-housing-in-washington-dc-and-beyond/) of Brookings in 2019:

>[DC's] zoning and politics insulate the city's wealthy areas from change, while lower-income families struggle to afford ever-increasing rents.

Here's [Lance Freeman](https://www.brookings.edu/articles/build-race-equity-into-rezoning-decisions/) of Columbia University in 2021 (emphasis added):

>Since the early 20th century, **zoning has continued to be a tool used to separate the poor, who are disproportionately people of color, from certain neighborhoods.** This can be done, for example, by **allowing only single-family homes or homes on large lots, both of which will exclude more affordable housing.**

And here's what DC's current [zoning handbook](https://handbook.dcoz.dc.gov/pages/residential-r-zones) says is the purpose of the R-1 zone (emphasis added):

>The R-1A zone is intended to provide areas predominantly developed with **detached houses on large lots.**

DC's current zoning handbook even uses verbiage that echoes the words of 20th century segregationists. R-1 zones are intended to "Protect quiet residential areas." R-2 zoning is supposed to "Protect these areas from invasion by denser types of residential development." 

For comparison, here's urbanist [Marc Weiss](https://escholarship.org/content/qt26b8d8zh/qt26b8d8zh_noSplash_38cfb754168650ea762c01485cf19053.pdf) describing segregationists' efforts in 20s-era Berkeley:

>California Real Estate magazine reported that one of the most popular examples of "service rendered" by a local real estate board was the attempt by the Berkeley Realty Board, a strong supporter of city planning and zoning, "to organize a district of some twenty blocks under the covenant plan as protection against invasion of Negroes and Asiatics."

<br><br>

**It's time to do what the experts recommend. Land-use policymakers should stop "protecting" neighborhoods from "invasion," and delete DC's R-1 and R-2 zones, because they're perpetuating socioeconomic segregation.**

<br><br><br><br>


Data sources: [Open Data DC](https://opendata.dc.gov/), [Mapping Segregation DC](https://www.mappingsegregationdc.org/download/residential-sub-areas-for-website-rev.pdf).

